---
title: "Empirical Copula → TrialSimulator: End-to-End Example (Rank Resampling)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical Copula to TrialSimulator (Rank Resampling)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "<style>\n",
  "pre.sourceCode { background-color: #f7f7f7; }\n",
  "pre > code.sourceCode > span { display: block; }\n",
  "</style>\n",
  sep = ""
)
```

This vignette demonstrates how to wrap the `simdatacopulas` empirical copula (empirical copula / rank resampling) joint generator into the `generator(n)` interface required by `TrialSimulator`, and how to:

1. Reproduce dependence using the joint rank structure (`u_data`) from pilot/historical data
2. Control marginal distributions using your chosen marginal quantile functions (`dist`)
3. Plug into TrialSimulator and run a full trial simulation workflow

Key characteristics of the empirical copula:

- Strength: directly reproduces the joint rank structure observed in historical data (no need to choose a parametric copula family)
- Limitation: weak extrapolation, especially for "unobserved extreme tail structure", which is hard to create out of thin air

## 1. Load Dependencies

```{r}
library(simdata)
library(simdatacopulas)
library(TrialSimulator)
library(copula)
```

## 2. Prepare Pilot Data and Extract `u_data`

The input to an empirical copula is typically `u_data`: pseudo-observations obtained by applying `pobs()` to historical samples, which essentially preserve the joint rank structure.

For reproducibility, we construct pilot data in this example; in your real project, you should replace `pilot` with real historical data.

```{r}
set.seed(20251230)

n_pilot <- 2000

Sigma <- matrix(c(
  1.0,  0.5,  0.3,  0.2,
  0.5,  1.0,  0.4,  0.3,
  0.3,  0.4,  1.0,  0.6,
  0.2,  0.3,  0.6,  1.0
), nrow = 4, byrow = TRUE)

L <- chol(Sigma)
E <- matrix(stats::rnorm(n_pilot * 4), nrow = n_pilot, ncol = 4)
Z <- E %*% t(L)
U <- stats::pnorm(Z)

pilot <- data.frame(
  PFS_raw     = stats::qlnorm(U[, 1], meanlog = log(8),  sdlog = 0.60),
  OS_extra    = stats::qlnorm(U[, 2], meanlog = log(6),  sdlog = 0.60),
  BIOM_latent = stats::qnorm(U[, 3],  mean = 0,         sd = 1.0),
  ORR_latent  = stats::qnorm(U[, 4],  mean = 0,         sd = 1.0)
)

u_data <- copula::pobs(as.matrix(pilot))
u_data <- pmin(pmax(u_data, 1e-6), 1 - 1e-6)
```

## 3. Define Marginals (dist) for Simulation

Typical usage of the empirical copula:

- Dependence: from historical `u_data`
- Marginals: from your specified `dist` (can differ from historical marginals)

Here we encode a treatment effect by using different marginal parameters for placebo vs treatment.

```{r}
dist_pbo <- list(
  function(u) stats::qlnorm(u, meanlog = log(8),  sdlog = 0.55),
  function(u) stats::qlnorm(u, meanlog = log(6),  sdlog = 0.55),
  function(u) stats::qnorm(u,  mean = 0,         sd = 1.0),
  function(u) stats::qnorm(u,  mean = 0,         sd = 1.0)
)

dist_trt <- list(
  function(u) stats::qlnorm(u, meanlog = log(10), sdlog = 0.55),
  function(u) stats::qlnorm(u, meanlog = log(8),  sdlog = 0.55),
  function(u) stats::qnorm(u,  mean = -0.3,       sd = 1.0),
  function(u) stats::qnorm(u,  mean = 0.5,        sd = 1.0)
)
```

## 4. Build the Empirical-Copula `simdesign`

`simdesign_empirical_copula()` resamples the joint rank structure from `u_data` (rank resampling), then maps each column to the target scale using `dist`.

```{r}
design_pbo <- simdesign_empirical_copula(
  u_data      = u_data,
  dist        = dist_pbo,
  names_final = c("PFS_raw", "OS_extra", "BIOM_latent", "ORR_latent"),
  replace     = TRUE,
  jitter      = 0
)

design_trt <- simdesign_empirical_copula(
  u_data      = u_data,
  dist        = dist_trt,
  names_final = c("PFS_raw", "OS_extra", "BIOM_latent", "ORR_latent"),
  replace     = TRUE,
  jitter      = 0
)
```

## 5. Wrap into TrialSimulator `generator(n)`

Key points of the TrialSimulator endpoint generator contract:

- The first argument name must be `n`
- Return a `data.frame` with `n` rows
- For TTE: must provide a `<name>_event` column
- For non-tte: must provide `readout`

This example constructs 4 endpoints:

- `PFS`, `OS`: TTE, with `OS = PFS + extra` ensuring `OS >= PFS`
- `ORR`: binary non-tte, obtained by thresholding `ORR_latent`
- `BIOM`: continuous non-tte

```{r}
make_empirical_endpoint_rng <- function(design, orr_cutoff) {
  function(n) {
    latent <- simdata::simulate_data(design, n_obs = n)

    pfs_raw  <- latent[["PFS_raw"]]
    os_extra <- latent[["OS_extra"]]

    PFS <- pfs_raw
    OS  <- pfs_raw + os_extra

    data.frame(
      PFS       = PFS,
      PFS_event = rep(1L, n),
      OS        = OS,
      OS_event  = rep(1L, n),
      ORR       = as.integer(latent[["ORR_latent"]] > orr_cutoff),
      BIOM      = latent[["BIOM_latent"]]
    )
  }
}

rng_pbo <- make_empirical_endpoint_rng(design_pbo, orr_cutoff = stats::qnorm(0.6))
rng_trt <- make_empirical_endpoint_rng(design_trt, orr_cutoff = stats::qnorm(0.4))

head(rng_pbo(n = 3))
head(rng_trt(n = 3))
```

## 6. TrialSimulator: Define endpoints/arms/trial/milestones and run

```{r}
enroller_fn <- function(n) stats::rexp(n, rate = log(2) / 4)
dropout_fn  <- function(n) stats::rexp(n, rate = log(2) / 40)

ep_pbo <- endpoint(
  name      = c("PFS", "OS", "ORR", "BIOM"),
  type      = c("tte", "tte", "non-tte", "non-tte"),
  readout   = c(ORR = 6, BIOM = 0),
  generator = rng_pbo
)

ep_trt <- endpoint(
  name      = c("PFS", "OS", "ORR", "BIOM"),
  type      = c("tte", "tte", "non-tte", "non-tte"),
  readout   = c(ORR = 6, BIOM = 0),
  generator = rng_trt
)

arm_pbo <- arm(name = "placebo")
arm_pbo$add_endpoints(ep_pbo)

arm_trt <- arm(name = "treatment")
arm_trt$add_endpoints(ep_trt)

trial_obj <- trial(
  name       = "empirical-copula-demo",
  n_patients = 400,
  duration   = 36,
  enroller   = enroller_fn,
  dropout    = dropout_fn,
  seed       = 20251230,
  silent     = TRUE
)
trial_obj$add_arms(sample_ratio = c(1, 1), arm_pbo, arm_trt)

milestone_interim <- milestone(
  name   = "interim",
  when   = eventNumber(endpoint = "PFS", n = 100),
  action = doNothing
)

milestone_final <- milestone(
  name   = "final",
  when   = calendarTime(time = 36),
  action = doNothing
)

listener_obj <- listener()
listener_obj$add_milestones(milestone_interim, milestone_final)

controller_obj <- controller(trial_obj, listener_obj)
controller_obj$run(plot_event = FALSE)
```

## 7. Read Locked Data and Run Basic Checks

```{r}
locked_interim <- trial_obj$get_locked_data("interim")
locked_final   <- trial_obj$get_locked_data("final")

head(locked_interim)
head(locked_final)

all(locked_final$OS >= locked_final$PFS, na.rm = TRUE)
with(locked_final, stats::cor(PFS, OS, use = "complete.obs"))

with(subset(locked_final, arm == "placebo"), mean(ORR, na.rm = TRUE))
with(subset(locked_final, arm == "treatment"), mean(ORR, na.rm = TRUE))

with(subset(locked_final, arm == "placebo"), mean(BIOM, na.rm = TRUE))
with(subset(locked_final, arm == "treatment"), mean(BIOM, na.rm = TRUE))

with(locked_final, tapply(PFS_event, arm, mean, na.rm = TRUE))
with(locked_final, tapply(OS_event,  arm, mean, na.rm = TRUE))

with(locked_final, tapply(is.na(ORR), arm, mean))
with(locked_final, tapply(is.na(BIOM), arm, mean))
```

## 8. Common Variants

- If your historical data have many ties (highly discretized), consider using `jitter > 0` to break ties.
- If you want to exactly reproduce historical marginals (instead of custom marginals), use `simdesign_empirical_copula_from_data(data, vars, ...)` to generate `u_data` directly from the data and use empirical quantiles as marginals.

