---
title: "Vine Copula → TrialSimulator: End-to-End Example (R-vine)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vine Copula to TrialSimulator (R-vine)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r, echo = FALSE, results = "asis"}
cat(
  "<style>\n",
  "pre.sourceCode { background-color: #f7f7f7; }\n",
  "pre > code.sourceCode > span { display: block; }\n",
  "</style>\n",
  sep = ""
)
```

This vignette demonstrates how to wrap the `simdatacopulas` vine copula (C-vine / D-vine / R-vine) joint generator into the `generator(n)` interface required by `TrialSimulator`, and how to:

1. Fit (or specify) a vine dependence structure using `VineCopula`
2. Control marginal distributions using your chosen marginal quantile functions (`dist`)
3. Plug into TrialSimulator and run milestones / database lock / read locked data

This example uses an R-vine (the most general and structurally flexible form).

## 1. Load Dependencies

```{r}
library(simdata)
library(simdatacopulas)
library(TrialSimulator)
```

This vignette also requires `VineCopula` and `copula`:

```{r}
library(VineCopula)
library(copula)
```

## 2. Build/Fit a Vine Dependence Structure

For reproducibility, we first generate pseudo-observations `u_pilot` from a 4-dimensional Gaussian copula, then fit an R-vine structure using `RVineStructureSelect()` (in your real project, you can replace `u_pilot` with pseudo-observations obtained from historical data via `pobs()`).

```{r}
set.seed(20251230)

d <- 4L
n_pilot <- 2000L

cop_true <- copula::normalCopula(param = 0.6, dim = d, dispstr = "ex")
u_pilot <- copula::rCopula(n_pilot, cop_true)
u_pilot <- pmin(pmax(u_pilot, 1e-6), 1 - 1e-6)

vine_fit <- VineCopula::RVineStructureSelect(
  data = u_pilot,
  familyset = 1:6,
  indeptest = TRUE,
  selectioncrit = "AIC",
  method = "mle"
)
```

## 3. Define Marginals (dist) and Build `simdesign`

We construct 4 correlated variables:

- `PFS_raw`: raw PFS time-to-event (positive)
- `OS_extra`: additional survival time after PFS (positive)
- `BIOM_latent`: continuous biomarker
- `ORR_latent`: latent normal variable, later thresholded into binary ORR

Placebo and treatment share the same vine dependence structure, but use different marginal parameters to represent a "treatment effect".

```{r}
dist_pbo <- list(
  function(u) stats::qlnorm(u, meanlog = log(8),  sdlog = 0.55),
  function(u) stats::qlnorm(u, meanlog = log(6),  sdlog = 0.55),
  function(u) stats::qnorm(u,  mean = 0,         sd = 1.0),
  function(u) stats::qnorm(u,  mean = 0,         sd = 1.0)
)

dist_trt <- list(
  function(u) stats::qlnorm(u, meanlog = log(10), sdlog = 0.55),
  function(u) stats::qlnorm(u, meanlog = log(8),  sdlog = 0.55),
  function(u) stats::qnorm(u,  mean = -0.3,       sd = 1.0),
  function(u) stats::qnorm(u,  mean = 0.5,        sd = 1.0)
)

design_pbo <- simdesign_vine_copula(
  vine        = vine_fit,
  dist        = dist_pbo,
  names_final = c("PFS_raw", "OS_extra", "BIOM_latent", "ORR_latent"),
  eps         = 1e-6
)

design_trt <- simdesign_vine_copula(
  vine        = vine_fit,
  dist        = dist_trt,
  names_final = c("PFS_raw", "OS_extra", "BIOM_latent", "ORR_latent"),
  eps         = 1e-6
)
```

## 4. Wrap into TrialSimulator `generator(n)`

Key points of the TrialSimulator endpoint generator contract:

- The first argument name must be `n`
- Return a `data.frame` with `n` rows
- For TTE: must provide a `<name>_event` column
- For non-tte: must provide `readout`

This example constructs 4 endpoints:

- `PFS`, `OS`: TTE, with `OS = PFS + extra` ensuring `OS >= PFS`
- `ORR`: binary non-tte, obtained by thresholding `ORR_latent`
- `BIOM`: continuous non-tte

```{r}
make_vine_endpoint_rng <- function(design, orr_cutoff) {
  function(n) {
    latent <- simdata::simulate_data(design, n_obs = n)

    pfs_raw  <- latent[["PFS_raw"]]
    os_extra <- latent[["OS_extra"]]

    PFS <- pfs_raw
    OS  <- pfs_raw + os_extra

    data.frame(
      PFS       = PFS,
      PFS_event = rep(1L, n),
      OS        = OS,
      OS_event  = rep(1L, n),
      ORR       = as.integer(latent[["ORR_latent"]] > orr_cutoff),
      BIOM      = latent[["BIOM_latent"]]
    )
  }
}

rng_pbo <- make_vine_endpoint_rng(design_pbo, orr_cutoff = stats::qnorm(0.6))
rng_trt <- make_vine_endpoint_rng(design_trt, orr_cutoff = stats::qnorm(0.4))

head(rng_pbo(n = 3))
head(rng_trt(n = 3))
```

## 5. TrialSimulator: Define endpoints/arms/trial/milestones and run

```{r}
enroller_fn <- function(n) stats::rexp(n, rate = log(2) / 4)
dropout_fn  <- function(n) stats::rexp(n, rate = log(2) / 40)

ep_pbo <- endpoint(
  name      = c("PFS", "OS", "ORR", "BIOM"),
  type      = c("tte", "tte", "non-tte", "non-tte"),
  readout   = c(ORR = 6, BIOM = 0),
  generator = rng_pbo
)

ep_trt <- endpoint(
  name      = c("PFS", "OS", "ORR", "BIOM"),
  type      = c("tte", "tte", "non-tte", "non-tte"),
  readout   = c(ORR = 6, BIOM = 0),
  generator = rng_trt
)

arm_pbo <- arm(name = "placebo")
arm_pbo$add_endpoints(ep_pbo)

arm_trt <- arm(name = "treatment")
arm_trt$add_endpoints(ep_trt)

trial_obj <- trial(
  name       = "vine-copula-demo",
  n_patients = 400,
  duration   = 36,
  enroller   = enroller_fn,
  dropout    = dropout_fn,
  seed       = 20251230,
  silent     = TRUE
)
trial_obj$add_arms(sample_ratio = c(1, 1), arm_pbo, arm_trt)

milestone_interim <- milestone(
  name   = "interim",
  when   = eventNumber(endpoint = "PFS", n = 100),
  action = doNothing
)

milestone_final <- milestone(
  name   = "final",
  when   = calendarTime(time = 36),
  action = doNothing
)

listener_obj <- listener()
listener_obj$add_milestones(milestone_interim, milestone_final)

controller_obj <- controller(trial_obj, listener_obj)
controller_obj$run(plot_event = FALSE)
```

## 6. Read Locked Data and Run Basic Checks

```{r}
locked_interim <- trial_obj$get_locked_data("interim")
locked_final   <- trial_obj$get_locked_data("final")

head(locked_interim)
head(locked_final)

all(locked_final$OS >= locked_final$PFS, na.rm = TRUE)
with(locked_final, stats::cor(PFS, OS, use = "complete.obs"))

with(subset(locked_final, arm == "placebo"), mean(ORR, na.rm = TRUE))
with(subset(locked_final, arm == "treatment"), mean(ORR, na.rm = TRUE))

with(subset(locked_final, arm == "placebo"), mean(BIOM, na.rm = TRUE))
with(subset(locked_final, arm == "treatment"), mean(BIOM, na.rm = TRUE))

with(locked_final, tapply(PFS_event, arm, mean, na.rm = TRUE))
with(locked_final, tapply(OS_event,  arm, mean, na.rm = TRUE))

with(locked_final, tapply(is.na(ORR), arm, mean))
with(locked_final, tapply(is.na(BIOM), arm, mean))
```
